Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-12 main.tex
typos and tiny changes

\newcommand{\gA}{\graph{A}}
\newcommand{\gC}{\graph{C}}
\newcommand{\gG}{\graph{G}}
\newcommand{\gH}{\graph{H}}

@@

First-order predicate logic is one of the cornerstones of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer sience. Since
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on its own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
that has been made in proof theory in general and in the area of
automated and interactive theorem provers in
particular\cite{DSAV:15lean,paulson:isabelle}\todo{find
references}\juihsuan{done}, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

In general, a \emph{decomposition theorem} provides normal forms of
proofs, separating subsets of infernence rules of a proof system. A
prominent example of a decomposion theorem is Herbrand's
theorem~\cite{herbrand:phd}, which allows a separation between the
proposotional part and the quantifier part in a first-order
proof~\cite{gentzen:35:II,brunnler:06:herbrand}. Through the advent of
deep inference, new kinds of proof decompositions became possible,
most notably the separation between the linear part of a proof and the
resource management of a proof. It has been shown by
Stra{\ss}burger~\cite{str:07:RTA} that a proof in classical
propositional logic can be decomposed into a proof of (multiplicative)
linear logic, followed by a proof consisting only of contractions and
weakenings. In this paper we show that the same is possible for
first-order logic.

Combinatorial proofs and deep infernence can be seen as opposite ends
of a spectrum: whereas deep inference allows for a very fine
granularity of inference rules---one inference rule in a standard
formalism, like sequent calculus or semantic tableaux, is usually
simulated by a sequence of different deep inference rules---have
combinatorial proofs completely abolished the concept of inference
rule. Nonetheless, there is a close relationship between the two,
realized through a decomposition theorem, as we establish it in this
paper.

@@

A graph homomorphism $f: (\vG, \eG) \rightarrow (\vH, \eH)$ is a \bfit{fibration} if
for all $v \in \vG$ and $wf(v) \in \eH$, there exists a unique $\tilde{w} \in
\vG$ such that $\tilde{w}v \in \eG$ and $f(\tilde{w}) = w$, and is a \bfit{skew fibration}
if for all $v \in \vG$ and $wf(v) \in \eH$ there exists $\tilde{w} \in \vG$ such that
$\tilde{w}v \in \eG$ and $f(\tilde{w})w \notin \eH$. A \bfit{fograph homomorphism}
is a graph homomorphism between the underlying graphs of the fographs
considered. A fograph homomorphism $f : \gG \rightarrow \gH$ is
\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
\bfit{existential-preserving} if for all existental binder $b$ in $\gG$, the
vertex $f(b)$ is an existential binder in $\gH$.

@@

\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $G$ is a skew
bifibration $f: N \rightarrow \gG$ where $N$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}

@@

Contrary to standard proof formalisms, like sequent calculi or
tableaux, where inference rules decompose the principal formula along
its root connective, can \emph{deep inference rules} be applied like
rewriting rules inside any (positve) formula or sequent
\bfit{context}, which is denoted as $\Sconhole$, and which is a
formula (resp.~sequent) with exactly one occurrence of the \bfit{hole}
$\conhole$ in the position of an atom. Then $\Scons A$ is the result
of replacing the hole $\conhole$ in $\Sconhole$ with $A$.

@@

\begin{thm}\label{thm:CP-DI}
  Let $A$ be a rectified formula, and assume we have a combinatorial proof $\phi\colon\gC\to\graphof A$. Then there is a derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wlD,\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  for some $A'$ such that $\graphof{\rectif{A'}}=\gC$. Conversely,
  whenever we have a derivation as in~\eqref{eq:decom} above, then
  there is a combinatiorial proof $\phi\colon\gC\to\graphof A$ such that $\gC=\graphof{\rectif{A'}}$.
\end{thm}

Furthermore, in the proof of Theorem~\ref{thm:CP-DI}, we will see that
(i) the links in the fonet $\gC$ correspond precicely to the pairs of
atoms that meet in the instances of the $\aiD$-rule in the derivation
$\Deri_1$, and (ii) the ''flow-graph'' of $\Deri_2$ that traces the
quatifier- and atom-occurrences in the derivation corresponds exactly
to the vertex-mapping induced by $\phi$.

Thus, combinatorial proofs are closely related to derivations of the
form~\eqref{eq:decom}, and since by Theorem~\ref{thm:decomposition}
every derivation can be transformed into that form, we can say that
combinatorial proofs form a canononical proof representation for first
order logic, similarly to what proof nets are for linear
logic~\cite{girard:96:PN}.

Finally, Theorems~\ref{thm:KS1}, \ref{thm:decomposition}
and~\ref{thm:CP-DI} imply Theorem~\ref{thm:FOCP}, which means that we
have here and alterantive proof of the soundness and competeness of
first-order combinatorial proofs which is simpler than the one given
in~\cite{hughes:fopws}.

@@

\subsection{Contraction and Weakining}

>>>>

\newcommand{\gA}{\graph{A}}
\newcommand{\gC}{\graph{C}}
\newcommand{\gG}{\graph{G}}
\newcommand{\gH}{\graph{H}}
\newcommand{\gN}{\graph{N}}

@@

First-order predicate logic is one of the cornerstones of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer science. Since
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on its own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
that has been made in proof theory in general and in the area of
automated and interactive theorem provers in
particular\cite{DSAV:15lean,paulson:isabelle}\todo{find
references}\juihsuan{done}, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

In general, a \emph{decomposition theorem} provides normal forms of
proofs, separating subsets of inference rules of a proof system. A
prominent example of a decomposition theorem is Herbrand's
theorem~\cite{herbrand:phd}, which allows a separation between the
propositional part and the quantifier part in a first-order
proof~\cite{gentzen:35:II,brunnler:06:herbrand}. Through the advent of
deep inference, new kinds of proof decompositions became possible,
most notably the separation between the linear part of a proof and the
resource management of a proof. It has been shown by
Stra{\ss}burger~\cite{str:07:RTA} that a proof in classical
propositional logic can be decomposed into a proof of (multiplicative)
linear logic, followed by a proof consisting only of contractions and
weakenings. In this paper we show that the same is possible for
first-order logic.

Combinatorial proofs and deep inference can be seen as opposite ends
of a spectrum: whereas deep inference allows for a very fine
granularity of inference rules---one inference rule in a standard
formalism, like sequent calculus or semantic tableaux, is usually
simulated by a sequence of different deep inference rules---have
combinatorial proofs completely abolished the concept of inference
rule. Nonetheless, there is a close relationship between the two,
realized through a decomposition theorem, as we establish it in this
paper.

@@

A graph homomorphism $f: (\vG, \eG) \rightarrow (\vH, \eH)$ is a \bfit{fibration} if
for all $v \in \vG$ and $wf(v) \in \eH$, there exists a unique $\tilde{w} \in
\vG$ such that $\tilde{w}v \in \eG$ and $f(\tilde{w}) = w$, and is a \bfit{skew fibration}
if for all $v \in \vG$ and $wf(v) \in \eH$ there exists $\tilde{w} \in \vG$ such that
$\tilde{w}v \in \eG$ and $f(\tilde{w})w \notin \eH$. A \bfit{fograph homomorphism}
is a graph homomorphism between the underlying graphs of the fographs
considered. A fograph homomorphism $f : \gG \rightarrow \gH$ is
\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
\bfit{existential-preserving} if for all existential binder $b$ in $\gG$, the
vertex $f(b)$ is an existential binder in $\gH$.

@@

\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $f: \gN \rightarrow \gG$ where $\gN$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}

@@

Contrary to standard proof formalisms, like sequent calculi or
tableaux, where inference rules decompose the principal formula along
its root connective, can \emph{deep inference rules} be applied like
rewriting rules inside any (positive) formula or sequent
\bfit{context}, which is denoted as $\Sconhole$, and which is a
formula (resp.~sequent) with exactly one occurrence of the \bfit{hole}
$\conhole$ in the position of an atom. Then $\Scons A$ is the result
of replacing the hole $\conhole$ in $\Sconhole$ with $A$.

@@

\begin{thm}\label{thm:CP-DI}
  Let $A$ be a rectified formula, and assume we have a combinatorial proof $\phi\colon\gC\to\graphof A$. Then there is a derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wlD,\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  for some $A'$ such that $\graphof{\rectif{A'}}=\gC$. Conversely,
  whenever we have a derivation as in~\eqref{eq:decom} above, then
  there is a combinatorial proof $\phi\colon\gC\to\graphof A$ such that $\gC=\graphof{\rectif{A'}}$.
\end{thm}

Furthermore, in the proof of Theorem~\ref{thm:CP-DI}, we will see that
(i) the links in the fonet $\gC$ correspond precisely to the pairs of
atoms that meet in the instances of the $\aiD$-rule in the derivation
$\Deri_1$, and (ii) the ''flow-graph'' of $\Deri_2$ that traces the
quantifier- and atom-occurrences in the derivation corresponds exactly
to the vertex-mapping induced by $\phi$.

Thus, combinatorial proofs are closely related to derivations of the
form~\eqref{eq:decom}, and since by Theorem~\ref{thm:decomposition}
every derivation can be transformed into that form, we can say that
combinatorial proofs form a canonical proof representation for first
order logic, similarly to what proof nets are for linear
logic~\cite{girard:96:PN}.

Finally, Theorems~\ref{thm:KS1}, \ref{thm:decomposition}
and~\ref{thm:CP-DI} imply Theorem~\ref{thm:FOCP}, which means that we
have here an alternative proof of the soundness and completeness of
first-order combinatorial proofs which is simpler than the one given
in~\cite{hughes:fopws}.

@@

\subsection{Contraction and Weakening}
